Supplementary Information

### 2021-06-03 doc/manuscript/ForestGEO_treerings_SI.Rmd
fix SI references
#131

In total, 2210 individuals were cored either within the 25 ha ForestGEO plot or a larger, 74 ha plot in which it is embedded. 
The tree-ring data were collected for several studies around 2010 [@samonil_individual-based_2013; @samonil_disturbances_2014; @samonil_uncertainty_2015; @samonil_uncertainty_2016; @vasickova_true_2016; @vasickova_driving_2019; @kaspar_species-specific_nodate]. 
The sampling strategy was mainly connected with the disturbance history of the forest and evaluation of growth of juvenile trees in gaps. 
Therefore, these data consist mainly of dominant and codominant tree species and relatively low number of young trees.

@@

### Cedar Breaks, Utah, USA
Cores were collected within the Utah Forest Dynamics Plot ForestGEO plot on the Colorado Plateau in Cedar Breaks National Monument [CB; @furniss_reconciling_2017]. 
The plot sits on the edge of cliff, which descends ~750 m to a valley below.
The mature spruce-fir forest is dominated by the species analyzed here (*Abies bifolia*, *Populus tremuloides*, *Picea engelmannii*, *Pinus flexilis*, *Pinus longaeva*, *Picea pungens*, and *Pseudotsuga menziesii*).
The site is largely undisturbed. 
The most recent fire occurred in 1802, and many currently living trees established before then. Living *Pinus longaeva* at the site date back ~1600 years, and deadwood at the site dates back to 0 AD [@lutz_large-diameter_2021].

In 2019, living individuals of each dominant species were cored.  
Chronologies are published on ITRDB [@birch_birch_2020; @birch_birch_2020-2; @birch_birch_2020-3; @birch_birch_2020-4].

@@

Additionally, we used 800 m resolution PRISM records [@daly_mapping_2015; @daly_physiographically_2008-1] from 1930 to 2015 to test and correct for cases where the coarser-resolution CRU values were not locally representative. 
Variables in the PRISM data set included $T_{max}$, $T_{mean}$, $T_{min}$, and PPT.
To test for substantial differences between CRU and PRISM records, each variable was compared on a monthly basis using linear regression. We considered CRU records to be inappropriate if (1) a paired t-test revealed a significant difference between CRU and PRISM records, and (2) the absolute value of the mean monthly difference between CRU and PRISM exceeded 2.5°C (temperature variables) or 10 mm per month (precipitation).
Two sites met these criteria: LT and CB, both located in mountainous terrain in the western United States. 
For these sites, we used linear regression to characterize the relationship between the CRU and PRISM, and then used this relationship to correct CRU data outside the time frame of PRISM (i.e., pre-1930 and post-2015). 
CRU records were replaced with PRISM records over the time frame for which they were available.
For these sites, CRU records for variables not included in the PRISM data set (PDF, PET) were considered unreliable and excluded from the analysis.

@@

The third approach involved splitting analyses into decades before and after 1970 [*sensu*  @sniderhan_growth_2016] -- an approach that has also been applied in other studies in rapidly warming climates [e.g., @lloyd_no_2002; @driscoll_divergent_2005;@wilmking_recent_2004].
This approach should accurately capture the most important climatic drivers of growth variation prior to the acceleration of climate change in recent decades, and should also identify the drivers of recent high-frequency growth variation.
For PPT at LT, the most influential seasonal windows pre-1970 and post-1970 were similar to those for the entire time period.
For the temperature variable at LT, $T_{max}$ was consistently identified as the most important variable, but it's most influential seasonal window extended back to the previous growing season pre-1970, but was limited to the current growing season post-1970 and for the analysis as a whole.
For the water variable at SC, PPT was consistently selected as a better predictor than PDF, and responses pre- and post-1970 were generally similar to those over the full time period.
For the temperature variable at SC, the pre- and post- 1970 time periods both identified $T_{min}$ temperatures during a single month in the previous summer as the most influential climate driver for RW, whereas the analysis for the full time period identified $T_{max}$ of current July (Fig. `r SC_1970break`).
For the water variable at SCBI, PPT was consistently selected as a better predictor than PDF, with similar seasonal windows pre-1970 and for the full analysis period, compared to only a weak effect over a different time period (current May) post-1970 (Fig. `r SC_1970break`).
For the temperature variable at SCBI, selected variables (PET, $T_{max}$, or $T_{mean}$) and seasonal windows were variable across the three analysis periods, with only weak relationships in the post-1970 analysis period.
These results indicate potentially shifting climate sensitivities, particularly at SC (Fig. `r SC_1970break`), which may not be adequately captured by our baseline method.
Further analysis into how climate sensitivities may be shifting would be valuable, but is beyond the scope of this analysis.

@@

\newpage
`r qualitative_results_comparison`, cont.
```{r eval = TRUE, echo=FALSE, warning=FALSE}
table_pt3 <- read.csv("tables_figures/qualitative_results_comparison.csv", 
                     colClasses=c("NULL", NA, NA, NA, NA, NA, NA), 
                     col.names=c("NULL", "species", "previously observed", "observed here", "previously observed", "observed here", "reference"), 
                     skip=breakrow2, nrows=breakrow3-breakrow2,
                     stringsAsFactors = FALSE, check.names = FALSE)
kable(table_pt3, format = "latex", booktabs = TRUE, longtable = TRUE,  escape = F, linesep = c('\\addlinespace'))%>%
  add_header_above(c(" ", "Precipitation response" = 2, "Temperature response" = 2)) %>%
  column_spec(2:5, width = col_width) %>%
  column_spec(6, width = "2cm") %>%
  pack_rows("Žofín Forest Dynamics Plot, Czech Republic", 1, 4, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Niobrara, Nebraska, USA", 5, 5, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Little Tesuque, New Mexico, USA", 6, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Cedar Breaks, Utah, USA", 9, 9, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Scotty Creek, NW Territories, Canada", 10, 10, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7) 
# refs in this table need to go in the nocite field below
```
---
nocite: | 
  @alfaro-sanchez_growth_2017; @vlam_temperature_2014; @helcoski_growing_2019; @maxwell_declining_2016; @bumann_assessing_2019; @tumajer_increasing_2017; @sniderhan_growth_2016; @kaspar_species-specific_nodate; @touchan_millennial_2011; @williams_temperature_2013; @alexander_potential_2019
---
*Indicates results from a regional study including but not limited to cores from the focal site.

>>>>

In total, 2210 individuals were cored either within the 25 ha ForestGEO plot or a larger, 74 ha plot in which it is embedded. 
The tree-ring data were collected for several studies around 2010 [@samonil_individualbased_2013; @samonil_disturbances_2014; @samonil_uncertainty_2015; @samonil_uncertainty_2016; @vasickova_true_2016; @vasickova_driving_2019; @kaspar_speciesspecific_2021]. 
The sampling strategy was mainly connected with the disturbance history of the forest and evaluation of growth of juvenile trees in gaps. 
Therefore, these data consist mainly of dominant and codominant tree species and relatively low number of young trees.

@@

### Cedar Breaks, Utah, USA
Cores were collected within the Utah Forest Dynamics Plot ForestGEO plot on the Colorado Plateau in Cedar Breaks National Monument [CB; @furniss_reconciling_2017]. 
The plot sits on the edge of cliff, which descends ~750 m to a valley below.
The mature spruce-fir forest is dominated by the species analyzed here (*Abies bifolia*, *Populus tremuloides*, *Picea engelmannii*, *Pinus flexilis*, *Pinus longaeva*, *Picea pungens*, and *Pseudotsuga menziesii*).
The site is largely undisturbed. 
The most recent fire occurred in 1802, and many currently living trees established before then. Living *Pinus longaeva* at the site date back ~1600 years, and deadwood at the site dates back to 0 AD [@lutz_largediameter_2021].

In 2019, living individuals of each dominant species were cored.  
Chronologies are published on ITRDB [@birch_birch_2020; @birch_birch_2020a; @birch_birch_2020b; @birch_birch_2020c].

@@

Additionally, we used 800 m resolution PRISM records [@daly_mapping_2015; @daly_physiographically_2008] from 1930 to 2015 to test and correct for cases where the coarser-resolution CRU values were not locally representative. 
Variables in the PRISM data set included $T_{max}$, $T_{mean}$, $T_{min}$, and PPT.
To test for substantial differences between CRU and PRISM records, each variable was compared on a monthly basis using linear regression. We considered CRU records to be inappropriate if (1) a paired t-test revealed a significant difference between CRU and PRISM records, and (2) the absolute value of the mean monthly difference between CRU and PRISM exceeded 2.5°C (temperature variables) or 10 mm per month (precipitation).
Two sites met these criteria: LT and CB, both located in mountainous terrain in the western United States. 
For these sites, we used linear regression to characterize the relationship between the CRU and PRISM, and then used this relationship to correct CRU data outside the time frame of PRISM (i.e., pre-1930 and post-2015). 
CRU records were replaced with PRISM records over the time frame for which they were available.
For these sites, CRU records for variables not included in the PRISM data set (PDF, PET) were considered unreliable and excluded from the analysis.

@@

The third approach involved splitting analyses into decades before and after 1970 [*sensu*  @sniderhan_growth_2016] -- an approach that has also been applied in other studies in rapidly warming climates [e.g., @lloyd_spatial_2002; @driscoll_divergent_2005;@wilmking_recent_2004].
This approach should accurately capture the most important climatic drivers of growth variation prior to the acceleration of climate change in recent decades, and should also identify the drivers of recent high-frequency growth variation.
For PPT at LT, the most influential seasonal windows pre-1970 and post-1970 were similar to those for the entire time period.
For the temperature variable at LT, $T_{max}$ was consistently identified as the most important variable, but it's most influential seasonal window extended back to the previous growing season pre-1970, but was limited to the current growing season post-1970 and for the analysis as a whole.
For the water variable at SC, PPT was consistently selected as a better predictor than PDF, and responses pre- and post-1970 were generally similar to those over the full time period.
For the temperature variable at SC, the pre- and post- 1970 time periods both identified $T_{min}$ temperatures during a single month in the previous summer as the most influential climate driver for RW, whereas the analysis for the full time period identified $T_{max}$ of current July (Fig. `r SC_1970break`).
For the water variable at SCBI, PPT was consistently selected as a better predictor than PDF, with similar seasonal windows pre-1970 and for the full analysis period, compared to only a weak effect over a different time period (current May) post-1970 (Fig. `r SC_1970break`).
For the temperature variable at SCBI, selected variables (PET, $T_{max}$, or $T_{mean}$) and seasonal windows were variable across the three analysis periods, with only weak relationships in the post-1970 analysis period.
These results indicate potentially shifting climate sensitivities, particularly at SC (Fig. `r SC_1970break`), which may not be adequately captured by our baseline method.
Further analysis into how climate sensitivities may be shifting would be valuable, but is beyond the scope of this analysis.

@@

\newpage
`r qualitative_results_comparison`, cont.
```{r eval = TRUE, echo=FALSE, warning=FALSE}
table_pt3 <- read.csv("tables_figures/qualitative_results_comparison.csv", 
                     colClasses=c("NULL", NA, NA, NA, NA, NA, NA), 
                     col.names=c("NULL", "species", "previously observed", "observed here", "previously observed", "observed here", "reference"), 
                     skip=breakrow2, nrows=breakrow3-breakrow2,
                     stringsAsFactors = FALSE, check.names = FALSE)
kable(table_pt3, format = "latex", booktabs = TRUE, longtable = TRUE,  escape = F, linesep = c('\\addlinespace'))%>%
  add_header_above(c(" ", "Precipitation response" = 2, "Temperature response" = 2)) %>%
  column_spec(2:5, width = col_width) %>%
  column_spec(6, width = "2cm") %>%
  pack_rows("Žofín Forest Dynamics Plot, Czech Republic", 1, 4, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Niobrara, Nebraska, USA", 5, 5, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Little Tesuque, New Mexico, USA", 6, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Cedar Breaks, Utah, USA", 9, 9, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Scotty Creek, NW Territories, Canada", 10, 10, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7) 
# refs in this table need to go in the nocite field below
```
---
nocite: | 
  @alfaro-sanchez_growth_2017; @vlam_temperature_2014; @helcoski_growing_2019; @maxwell_declining_2016; @bumann_assessing_2019; @tumajer_increasing_2017; @sniderhan_growth_2016; @kaspar_speciesspecific_2021; @touchan_millennial_2011; @williams_temperature_2013; @alexander_potential_2019
---
*Indicates results from a regional study including but not limited to cores from the focal site.
